---
title: "Bellabeat Smart Device Analysis Report"
author: "Julien LAY"
date: "2026-02-16"
output: pdf_document
---

## Project Overview

This report presents a behavioral analysis of non-Bellabeat smart device usage to extract insights relevant for Bellabeat Leaf marketing and product positioning.

---

## Data Loading

```{r data-loading}
library(tidyverse)
library(stats)

data <- read_csv("../data/processed/final_dataset_prepared.csv")
```

---

## Descriptive Statistics

### Global Metrics

```{r summary-stats}
summary_stats <- data %>%
  summarise(
    mean_steps = mean(TotalSteps, na.rm = TRUE),
    median_steps = median(TotalSteps, na.rm = TRUE),
    sd_steps = sd(TotalSteps, na.rm = TRUE),
    mean_active = mean(total_active_minutes, na.rm = TRUE),
    mean_sedentary_ratio = mean(sedentary_ratio, na.rm = TRUE),
    mean_sleep = mean(TotalMinutesAsleep, na.rm = TRUE),
    mean_time_in_bed = mean(TotalTimeInBed, na.rm = TRUE),
    mean_calories = mean(Calories, na.rm = TRUE)
  )
summary_stats
```

### Weekday vs Weekend Comparison

```{r weekday-analysis}
weekday_analysis <- data %>%
  group_by(is_weekend) %>%
  summarise(
    mean_steps = mean(TotalSteps, na.rm = TRUE),
    mean_sleep = mean(TotalMinutesAsleep, na.rm = TRUE),
    mean_sedentary = mean(sedentary_ratio, na.rm = TRUE)
  )
weekday_analysis
```

---

## Behavioral Patterns

### Average Steps by Weekday

```{r avg-steps-weekday}
avg_steps_weekday <- data %>%
  group_by(weekday) %>%
  summarise(mean_steps = mean(TotalSteps, na.rm = TRUE)) %>%
  arrange(match(weekday, c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday")))

p_steps_weekday <- ggplot(avg_steps_weekday, aes(x = weekday, y = mean_steps)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  ggtitle("Average Steps by Weekday") +
  xlab("Weekday") +
  ylab("Average Steps")
p_steps_weekday
```

### Average Sleep by Weekday

```{r avg-sleep-weekday}
avg_sleep_weekday <- data %>%
  group_by(weekday) %>%
  summarise(mean_sleep_minutes = mean(TotalMinutesAsleep, na.rm = TRUE)) %>%
  arrange(match(weekday, c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday")))

p_sleep_weekday <- ggplot(avg_sleep_weekday, aes(x = weekday, y = mean_sleep_minutes)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  ggtitle("Average Sleep Minutes by Weekday") +
  xlab("Weekday") +
  ylab("Average Sleep (minutes)")
p_sleep_weekday
```

---

## Correlation Analysis

### Sleep vs Steps per User

```{r sleep-steps-corr}
user_corr <- data %>%
  group_by(Id) %>%
  summarise(
    cor_steps_sleep = if(sum(!is.na(TotalSteps) & !is.na(TotalMinutesAsleep)) > 1) {
      cor(TotalSteps, TotalMinutesAsleep, use = "complete.obs")
    } else {
      NA_real_
    }
  )
user_corr
```

---

## Behavioral Segmentation

### K-means Clustering

```{r kmeans-segmentation}
segmentation_vars <- data %>%
  group_by(Id) %>%
  summarise(
    avg_steps = mean(TotalSteps, na.rm = TRUE),
    avg_sleep = mean(TotalMinutesAsleep, na.rm = TRUE),
    avg_sedentary = mean(sedentary_ratio, na.rm = TRUE)
  ) %>%
  filter(!is.na(avg_steps) & !is.na(avg_sleep) & !is.na(avg_sedentary))

seg_scaled <- scale(segmentation_vars[, -1])
set.seed(123)
k_clusters <- kmeans(seg_scaled, centers = 3)
segmentation_vars$cluster <- k_clusters$cluster
segmentation_vars
```

### Clusters Visualization

```{r clusters-plot, echo=FALSE}
ggplot(segmentation_vars, aes(x = avg_steps, y = avg_sleep, color = factor(cluster))) +
  geom_point(size = 3) +
  ggtitle("User Segmentation: Steps vs Sleep") +
  xlab("Average Steps") +
  ylab("Average Sleep (minutes)") +
  scale_color_discrete(name = "Cluster")
```

---

## Conclusions

- Users show clear weekday/weekend variations in steps and sleep.  
- Clustering reveals distinct behavioral segments based on activity and sleep patterns.  
- These insights can inform marketing strategies and product positioning for Bellabeat Leaf.